# Introduction {#intro}

## Definitions and sources of information

Developing emissions inventories is a complex task. Therefore, before entering 
into the details of the functions, it is good to provide some initial definitions.

A good starting point is the book _"The Art of Emissions Inventorying"_ [@pulles2010art].
This book provides general description of what is an emissions inventory. One of the early
sources of information about emissions inventories and emissions factors is the
Environmental Protection Agency (U.S.EPA) and its compilation of emission factors AP42 (https://www.epa.gov/air-emissions-factors-and-quantification).
A third source of information are the European EMEP/EEA air pollutant emission inventory guidebook (https://www.eea.europa.eu/themes/air/emep-eea-air-pollutant-emission-inventory-guidebook/).

An emissions inventory is the compilation of all mass emitted into the atmosphere
by activities in a defined area during a lapse of time. The Eq. (\@ref(eq:efGERAL) shows the
general form [@guia]:

\begin{equation}
Emission_{pollutant}=\sum_{activities} Activity reate \cdot Emission factor_{activity, pollutant}
(\#eq:efGERAL)
\end{equation}

There are two main types of inventories based on the application, one type is for policy
application and the other for scientific application [@pulles2010art]. 

- Inventories for policy application include the National Greenhouse Gas Inventory for the parties under the United Nations Framework on Climate Change [@unfccc].
- Emissions inventories with scientific applications include the Emission 
Database for Global Atmospheric Research EDGAR [@edgar].

Emissions inventories can be also multimedia inventories, such as the Pollutant Release
and Transfer Registers (PRTR) which include pollutants released to air, water and soil 
and transferred off-site for treatment or disposal (http://www.oecd.org/chemicalsafety/pollutant-release-transfer-register/). However,
the type of inventory covered in this book cover only the emissions released into the 
atmosphere.

### Approaches

It is necessary show some definitions of vehicular emissions inventories approaches, 
which in this case comes from the European Emissions Guidelines [@NtziachristosSamaras2016]:

- **Top-down** inventories are using input activity traffic data as fleet statistics,
fuel consumption, representative speeds and country balances. The emissions factors
are based on representative speeds. These inventories are also know as _static_.
- **Bottom-up** invetories are using input activity traffic data from traffic counts, traffic
simulations, vehicle composition, speed recordings and length of roads. The
emission factors are speed or/and acceleration functions. These inventories are also 
know as _dynamic_.
- **Reconciliation** Both approaches can be reconciliated in urban emissions inventories
with comparison of total mileage, cold start mileage or emission factors to ensure
that the comparison with the total fuel consumption in the study area is satisfactory.
This means that bottom-up and top-down vehicular emissions must match the fuel sales
over a region of study. 

## Installation

The VEIN r-package can be installed from Comprehensive R Archive Network 
[CRAN](https://cran.r-project.org/) directly:

```{r, eval=FALSE, echo = TRUE}
# CRAN
install.packages("vein")
```

VEIN can be also installed from github:

```{r, eval=FALSE}
library(devtools)
install_github("ibarraespinosa/vein")
```

The github installation requires that the _devtools_ package [@devtools] must be installed, including
its dependencies.

## Required R packages and dependencies

In order to use the VEIN package it is necesary to download other r-packages and its dependencies.
The current dependencies of VEIN are:

- **sp**. This is a package that provides classes and methods for spatial data [@sp] considering
points, lines, polygons and grids.
- **rgeos**. It is an interface to the "geometry Engine - Open Source ('GEOS') using the C 'API' for 
topology operations on geometries" [@rgeos]. It depends on the library GEOS (http://trac.osgeo.org/geos). 
- **rgdal**. Provides bindings to the 'Geospatial' Data Abstraction Library ('GDAL') (>= 1.6.3) and 
access to projection/transformation operations from the 'PROJ.4' library [@rgdal]. Depends on the library GDAL
(http://www.gdal.org/) and PROJ (http://proj4.org/).
- **raster**. It is a package that provides tools for analyses and modeling of geographic data [@raster].
- **units**. It is a package for measurement units in R [@units]. This package depends on the library 
udunits2 (http://www.unidata.ucar.edu/software/udunits/). I


```{r, eval=FALSE}
install.packages(c("sp", "rgeos", "rgdal", "raster", "units"))
```


In order to install these package it is necessary to install the dependencies into each operational system.
The instructions are:

**Ubuntu**
```{bash eval=FALSE}
sudo add-apt-repository ppa:ubuntugis/ubuntugis-unstable
sudo apt-get update
sudo apt-get install libudunits2-dev libgdal-dev libgeos-dev l
ibproj-dev 
```

**Fedora**
```{bash, eval=FALSE}
sudo dnf install gdal-devel proj-devel proj-epsg proj-nad 
geos-devel udunits2-devel
```

**Debian and other**

Dockerfiles from [rocker geospatial](https://github.com/rocker-org/geospatial).

**Windows**
Via [Rtools](https://cran.r-project.org/bin/windows/Rtools/).

**MAC OS**
```{bash, eval=FALSE}
brew unlink gdal
brew tap osgeo/osgeo4mac && brew tap --repair
brew install proj
brew install geos
brew install udunits
brew install gdal2 --with-armadillo --with-complete 
--with-libkml --with-unsupported
brew link --force gdal2
```

The new version of VEIN will replace the spatial r-packages dependencies of 
raster, rgeos and rgdal by the new **sf** [@sf] package and the **units** [@units] package.

**sf** is a package that provides simple features access for R [@sf]. It can be installed
in this way:

```{r, eval=FALSE}
install.packages("sf")
```

or via devtools

```{r eval=FALSE}
library(devtools)
install_github("r-spatial/sf")
```


## Data inside the VEIN model

There are several datasets available inside the model:

### Emission factors from CETESB

- **fe2015**: Emission factors from the Environmental Agency of Sao Paulo (CETESB).
Pollutants included: "CH4", "CO", "CO2", "HC", "N2O", "NMHC", "NOx" and "PM". The
type of vehicles included are Passenger Cars (PC).

usage:
```{r}
library(vein, quietly = T)
data(fe2015)
```

```{r, echo=FALSE}
knitr::kable(
    head(fe2015[, c(1:3, 11, 12)], 6), 
  caption = 'Emission factors from CETESB in VEIN model',
  booktabs = TRUE
)
```


### Mileage functions of Brazilian Fleet

- **fkm**: A list of functions of mileage in km for the Brazilian fleet. It includes mileage
functions based on more than 2 million recordings of vehicles [@BruniBales2013]. It consists
a list of age functions. The type of vehicles are Passneger Cars (PC), Light 
Commercial Vehicles (LCV), Small Busses (SBUS), Trucks, Articulated Trucks (ATRUCKS),
Motorcycles (MOTO) and Light vehicles (LDV). The fuels are Gasoline using 25\% of
ethanol (E25), Ethanol 100\% (E100) and Diesel with 5\% of biodiesel (B5). There are
also vehicles with flex engines which can run either with E25, E100 or any mixture
in between [@2005-01-4130]. The Fig. (\@ref(fig:mil) shows the mileage of PC using fuel E25.


usage:
```{r mil, fig.cap='Mileage of PC using E25', out.width='80%', fig.asp=.75, fig.align='center'}
library(vein, quietly = T)
data(fkm)
names(fkm)
age <- 1:50
mileage <- fkm[["KM_PC_E25"]](1:50) 
par(mar = c(4, 4, .1, .1))
plot(y = mileage, x = age, pch = 19, type = "b")
```


- **net**: Road network of the west part of Sao Paulo city. It consistes in a 
SpatialLinesDataFrame (class of sp) with the attributes ldv (light duty vehicles, 
$1 \cdot h^{-1}$), hdv (heavy duty vehicles, $1 \cdot h^{-1}$), ps (peak speed, 
$km \cdot h^{-1}$), ffs (free flow speed, $km \cdot h^{-1}$), tstreet (type of street), 
lanes (number of lanes), capacity (capacity fo vehicles at each link, 1/h) and tmin (time
for travelling each link, min). The Fig. (\@ref(traffic) shows more details.


### Temporal factors for Passenger Cars

Temporal factors (TF) are marix of jourly traffic data normalized for the hour
of interest, normally the morning rush hour 08:00-09:00 in local time (LT) [@gmd-2017-193].
This dataset cover 168 hours of one week and it is based on traffic counts
of toll stations near the center of the city of SãoPaulo, Brazil. The Fig.
(\@ref(fig:tf) shows the temporal factors for PC.

usage:
```{r tf, fig.cap='Temporal Factors for PC', out.width='80%', fig.asp=.75, fig.align='center'}
library(vein, quietly = T)
data(pc_profile)
tf <- unlist(pc_profile)
hours <- 1:168
par(mar = c(4, 4, .1, .1))
plot(y = tf, x = hours, pch = 16, type = "b")
```


### Profile for cold starts of Passenger Cars

This is a profile with the hourly percentage of cold starts of Passenger Cars. This
data covers 24 hours of a working day and it is based cold start recordings during 
the implementation of the International Vehicle Emissions (IVE) model [@Davisetal2005]
in São Paulo [@ivesp]. The Fig. (\@ref(fig:cold) shows the cold-start profile.

usage:
```{r cold, fig.cap='Cold-starts profile for PC', out.width='80%', fig.asp=.75, fig.align='center'}
library(vein, quietly = T)
data(pc_cold)
cold <- unlist(pc_cold)
hours <- 1:24
par(mar = c(4, 4, .1, .1))
plot(y = cold, x = hours, pch = 16, type = "b")
```

## Acknowledgements

Martin Ramacher, PhD Student Chemistry Transport Modelling, Institute for Coastal Research, Helmholtz-Zentrum Geesthacht, Germany - Proofreading.
Leila Droprinchinski Martins, Professor (Associate) Federal Technological University of Parana, Brazil - Comments.
