# Basics of R {#basic}

## Brief history

R is a programming language developed by Ross Ihaka and Robert Genleman in 1993 
[@ihaka1996r]. A good resource about R is the book  programming for data science 
[@peng2015r]. R can be explained as a free version of S-PLUS programming 
language (http://www.solutionmetrics.com.au/products/splus/default.html). It is 
free and recognized a GNU spftware (https://www.gnu.org/software/software.html)
with GNU General Public License (GPLV2-GPLV3) license. This feature allowed that many 
developers started improving and adding code over the years. 

R includes several statistical functions, therefore, user who just want to use 
the base capabilities of R are not forced to learn R-programming. However, the evolution from user to developer in R has been facilitated with numerous
publications such as the book "R packages: organize, test, document, and share
your code" [@wickham2015r], or "The art of R programming: A tour of statistical software  design" [@matloff2011art]. 

### Installation

R can be installed in different OS including Linux, Windows, Mac and Solaris. 
The webpage https://cran.r-project.org/ shows how to install R in any platform.
A popular Integrated development environment (IDE) is **RStudio** https://www.rstudio.com/, which contains many useful integrated options. 
However, R can be run on the terminal and use any text editor to write scripts.

## Using R

If we type at the R promt 
```{r}
x <- 1   # Not printing  results
x        # Print results
print(x) # Explicit print x
```

Rhe integer 1 was _assigned_ (<-) to x, then writing x or print(x) will show
the value of x. Instead of numbers, can be assigned other expressions
such as strings, dates or other objects. R includes several statistical 
functions. For instance, if we type 'pi' at the terminal, it will print
the pi number.

```{r}
pi
```

### R objects

There are five basic classes (or atomic classes).

- character.
- numeric.
- integer.
- complex.
- logical (TRUE/FALSE or just T or F).

Objects of the same class can be grouped in **vectors**. Vectors are created with 
writting **c()** with the elements of the vector inside the parenthesis, separated
by colon. In fact, **c** is a builtin function to create vectors. In this case,
the vector v1 contain a sequence of three integers, from 1 to 3. The resulting 
class is numeric.

```{r}
v1 <- 1:3
v2 <- c(1, 2, 3)
identical(v1, v2)
v1
class(v1)
```

With the operator `[` we can get specific elements of your vectors. In the 
following code it is also used the function `length`, which simply returns
the length of an object.

```{r}
v1[1]          # first element of v1
v1[length(v1)] # last element element of v1
```


If we create a vector with numbers and letter, the resulting vector will
be "character" converting the numbers in characters.

```{r}
v2 <- c(1:3,"a")
v2
class(v2)
```

Objects can be converted to other classes with the "as.*" functions. For instance,
If we convert the vector v2 to numeric, it will recognize the numbers adding an 
`r NA` into the position of the character.

```{r}
as.numeric(v2)
```


### Factors

Factors represents categorical data. A typical example is a character vector of
days of the week. The function `factor` creates factors, to see help type 
`?factor`. The following example will use this function with the arguments `x`
and `levels`. 

```{r}
dow <- factor(x = c("Wednesday", "Monday", "Thursday", "Friday"),
              levels =  c("Monday", "Tuesday", "Wednesday", "Thursday",
                          "Friday"))
dow
```

### Missing Values

Missing values in data-sets are a typical source of headache in r users. 
Fortunatly, R counts with several tools to avoid these headaches. These tools
are fhe functions `is.na` to check for NA (not available) and `is.nan` (not
a number). This functions returns logical values.

```{r}
n <- c(1, NA)
is.na(n)
n <- c(1, NaN)
is.na(n)
is.nan(n)
```


### Matrices

Matrices are structure with rows and columns. They can be created using the 
`matrix` function and also, using vectors. Remember, if you want to know more
about any function you just have to type ? and function, for instance:`?matrix`
will open the help documentation for the function `matrix`. Let's create a
matrix using vectors:

```{r}
a <- 1:12  #numeric vector
(m <- matrix(data = a, nrow = 3, ncol = 4))
```

We can check the dimensions of our matrix `m` with `dim`, which are 3 and 4.

```{r}
dim(m)
```

In order to get the elements of matrix `m` we can use `[` operator:
```{r}
m[1, 1] # first element
m[, 1]  # firs column
m[1, ]  # firs row
m[3,4]  # last element
```


### Arrays

Arrays are like matrices inside other matrices. In fact, a matrix is a 
2-dimensional array. Let's create an array create an array using the same vector
a and same dimensiones of m, 3 and 4. `array` has three arguments, `data`, `dim`
 and `dimnames`. In th argument `dim` let's add the number 2, resulting
 in an array of two matrices identical to `m`.

```{r}
(a <- array(data = a, dim = c(3,4,2)))
```


We can subset elements from array also with the `[` operator.

```{r}
a[1, 1, 1] # first element
dim(a)     # dimensions
a[3, 4, 2] # last element
```

### Lists

List are objects that can contain elements of different classes. I like to call
lists as bags. In a bag you can put almost anything. Also, inside the bag,
you can put other bags with different things. This means that you can have a
list of lists of any R object. Lists can be created with the `list` function.
Let's create an empty list. Then we will use the vecctor `a` to create another
list.

```{r}
a <- 1:3          # vector of three elements
l1 <- list()      # empty list
l1 <- list(a)     # vector of three elements
length(l1)        # length 1
l1 <- as.list(a)  # vector a as list
length(l1)        # three elements 
```

As mentioned, the list can have lists inside.

```{r}
l1 <- list(list(a),                    # three numeric elements
            list(TRUE, TRUE, FALSE))   # three logical elements
length(l1)                             # length 2
```

### Data-Frames

"Data-frames are used to store tabular data in R" [@peng2015r]. You can think
in data-frames as spreadsheet-like objects. They are similar with matrix and
you can have a matrix and a data-frame with the same dimensions. Basically,
you have rows and columns, columns can have different classes, and the columns
usually have a one-word name.

As this type of object is very used by scientists, there are R packages created
to work with this type of ojects. Below some of them:

- `data.frame`: This is not a package, for a class and function from the `base`.
With this functionyou can create data-frames.
- `data-table` [@datatable]: `data.tables` objects has the class `data.table` 
which inherits from the class `data-frame`, which means that they share common 
characteristics. However, the big difference with `data-table` is the efficiency
with memory. It includes function such as `fread` for reading millions of 
observations in seconds.
- `tidyr`[@tidyr]: Sometimes your data is in long format and you need it in wide
format, and viceversa. This package does the job.
- `readxl` [@readxl]: Good package for importing Excel and LibreOffice files. I
recommend  this packages for newbies.
- `sf` [@sf]: This package presents the class `sf` and introduce the list-column
of spatial geometry.

Let's create a data-frame.

```{r}
a <- 1:3
b <- 3:5
(ab <- data.frame(a, b))
class(ab)
```


### Names

In R you can put names in almost everything. Below examples for `vectors`, 
`matrix` and `data.frames`.

```{r}
names(ab)                # original names
names(ab) <- c("c", "d")
names(ab)                # new names
```

## Inputting data into R

Your data can be in different formats. In this sections i will show you how you
can input text and comma separated values files, which is a very common task. 
Let's say that you uses an spreadsheet program such as Microsoft Excel or
LibreOffice. Then you might want to export your data into R.

- Your first check if the first row of your data has name or not. If they do
have names, this means that they have a header.
- Then click on 'Save as' and search text file with extension .txt or CSV.
- Then you your file with a note bloc, gedit  or other tool to check if the
decimal character is point '.' and the separator character is comma ',' or 
semicolon ';' or other character. This depends on the regional confugarion of 
your spreadsheet software.
- If your file has extension '.txt', use `read.table` and if is '.csv', use
`read.csv`.

Once you check, you can import your '.txt' or '.csv' files in R:

```{r eval = FALSE}
a <- read.csv("path/to/file.csv", sep = ",", header = T)
a <- read.tablesv("path/to/file.txt", sep = ",", header = T)
```

