# (APPENDIX) Appendix {-} 

# Appendix A: Fast Example of VEIN  {-}


```{r eval = FALSE}
library(vein)
inventory(name = file.path(tempdir(), "YourCity"),
          show.dir = TRUE,
          show.scripts = TRUE)
source(paste0(file.path(tempdir(), "YourCity"), "main.R"))
```


# Appendix B: Detailed Example of VEIN  {-}
## Network


```{r eval = FALSE}
inventory(file.path(tempdir(), "YourCity"))
setwd("YourCity")
data(net) # Loads the traffic demand simulation for west São Paulo
data(net)
data(pc_profile)
pc_week <- temp_fact(net$ldv+net$hdv, pc_profile)
speed <- netspeed(pc_week, net$ps, net$ffs, net$capacity, net$lkm)
saveRDS(net, file = "network/net.rds")
saveRDS(speed, file = "network/speed.rds")
```

## Vehicular composition [@CETESB2015]

The vehicular composition `vehcomp` has 5 types of vehicles: Passenger Carss (PC),
Light Commercial Vehicles (LCV), Heavy Good Vehicles or trucks (HGV), Buses (BUS),
and Motorcycles (MC). The default value of this argument is:
`c(PC = 4, LCV = 5, HGV = 5, BUS = 3, MC = 9)`, which means that there are 4
types of PC, 5 of LCV, 5 of trucks, 3 of buses and 9 types of motorcycles. This
composition comes from the vehicular emissions inventory of the Environmental
Agency of São Paulo, Brazil [@CETESB2015]. In Brazil, the fuel used in vehicles
is blended with ethanol with and biodiesel. The user can use **any** vehicular
composition that representes its fleet with up-to 99 type of vehicles per
category. The default composition according Brazilian emissions inventory
is:

_Passenger Cars (PC)_

1. Passenger Cars using Gasoline blended with 25\% of ethanol (E25).
1. Passenger Cars with Flex engine using Gasoline blended with 25\% of ethanol (FE25).
1. Passenger Cars with Flex engine using 100\% of ethanol (FE100).
1. Passenger Cars with engines that uses only ethanol (E100).


_Light Commercial Vehicles (LCV)_

1. Light Commercial Vehicles with gross weight (GW) less than 3.8 t using
Gasoline blended with 25\% of ethanol (E25).
1. Light Commercial Vehicles with GW less than 3.8 t with Flex engine using
Gasoline blended with 25\% of ethanol (FE25).
1. Light Commercial Vehicles  with GW less than 3.8 t with Flex engine using
100\% of ethanol (FE100).
1. Light Commercial Vehicles  with GW less than 3.8 t with engines that uses
only ethanol (E100).
1. Light Commercial Vehicles  with GW less than 3.8 t with engines that uses
only ethanol (E100).

_Heavy Good Vehicles (HGV) or Trucks_

1. Semi Light Trucks (SLT) with 3.8 t < GW < 6 t using Diesel blended with 5\%
of biodiesel (B5).
1. Light Trucks (LT) with 6 t < GW < 10 t using Diesel blended with 5\% of
biodiesel (B5).
1. Medium Trucks (MT) with 10 t < GW < 15 t using Diesel blended with 5\% of
biodiesel (B5).
1. Semi Heavy Trucks (SHT) with 15 t < GW on a Rigid Truck (RT)  and GW < 40 on
an Articulated Truck (AT) using Diesel blended with 5\% of bodiesel (B5).
1. Heavy Trucks (HT) with 15 t < GW on a RT  and GW >= 40 on an AT using Diesel
blended with 5\% of bodiesel (B5).


_Buses_

1. Urban Bus (UB) using Diesel blended with 5\% of biodiesel (B5).
1. Small Urban Bus (SUB) using Diesel blended with 5\% of biodiesel (B5).
1. Motorway Bus (MB) or Coach using Diesel blended with 5\% of biodiesel (B5).

_Motorcycles (MC)_

1. Motorcycles with engine cc < 150 using Gasoline blended with 25\% of ethanol
(E25).
1. Motorcycles with engine 150 < cc < 500 using Gasoline blended with 25\% of 
ethanol (E25).
1. Motorcycles with engine cc > 500 using Gasoline blended with 25\% of ethanol
(E25).
1. Motorcycles with engine cc < 150 with Flex engine using Gasoline blended
with 25\% of ethanol (FE25).
1. Motorcycles with engine 150 < cc < 500 with Flex engine using Gasoline
blended with 25\% of ethanol (FE25).
1. Motorcycles with engine cc > 500 with Flex engine using Gasoline blended
with 25\% of ethanol (FE25).
1. Motorcycles with engine cc < 150 with Flex engine using 100\% of ethanol
(FE100).
1. Motorcycles with engine 150 < cc < 500 with Flex engine using 100\% of
ethanol (FE100).
1. Motorcycles with engine cc > 500 with Flex engine using 100\% of ethanol
(FE100).

## Traffic data

The vehicular composition will split the traffic simulation as shown on next
table. The simulation has traffic for Light Duty Vehicles and Heavy Good
Vehicles, therefore, the vehicular composition must split this categories and
considerations:

- Passenger Cars (PC) = 75% of Light Duty Vehicles (LDV) of traffic simulation.
- Light Commercial Vehicles (LCV) = 10% of LDV of traffic simulation.
- Motorcycles (MC) = 15% of LDV on traffic simulation.
- Heavy Good Vehicles (HGV) = 75% of Heavy Duty Vehicles (HDV).
- Buses =  = 25% of Heavy Duty Vehicles (HDV).
- LCV and PC have vehicles with flex engines capable to run with any mixture of
gasoline and ethanol [@2005-01-4130].
- Type of fuel consumed consists in gasoline blended with 25% of ethanol (E25).
- The vehicular composition consists in type of vehicles and type of fuel.
Optionally, the user could have a wider vehicular composition considering more
sizes, gross weight, etc.
- The percentages in composition apply for each type of vehicle PC, LCV, HGV,
BUS and MC.

**Passenger Cars**

The vehicular composition consisted in PC using E25, PC with Flex engines using
E25 and E100, and OC with engines for consuming only E100. The PC with Flex
engines entered into Brazilian market in 2003, therefore, at year 2015 flex
vehicles has 13 years of use. On the other case, PC with engines for E100 went
out of the market in 2007, therefore, the newest PC with E100 engine has 9 years
of use.

```{r comppc, echo=FALSE}
  Vehicles = c("PC_E25","PC_FE25", "PC_FE100", "PC_E100")
  Composition = c("37.25%","22.26%","37.97%", "2.44%")
df <- data.frame(Vehicles, Composition)
knitr::kable(df, 
  caption = 'Vehicular composition of PC for applied in this book',
  booktabs = TRUE,
  align = "c"
)
```


```{r eval = FALSE}
PC_E25 <- age_ldv(x = net$ldv,name = "PC_E25",
                  k = 75/100*37.25)
PC_FE25 <- age_ldv(x = net$ldv, name = "PC_FE25", agemax = 13,
                   k = 75/100*22.26)
PC_FE100 <- age_ldv(x = net$ldv, name = "PC_FE100", agemax = 13,
                    k = 75/100*37.97)
PC_E100 <- age_ldv(x = net$ldv, name = "PC_E100", agemin = 9,
                   k = 75/100*2.44)
saveRDS(PC_E25, file = "veh/PC_E25.rds")
saveRDS(PC_FE25, file = "veh/PC_FE25.rds")
saveRDS(PC_FE100, file = "veh/PC_FE100.rds")
saveRDS(PC_E100, file = "veh/PC_E100.rds")
```


**Light Commercial Vehicles**

The engine/fuel type affects in the same way to LCV with PC vehicles. However,
this category has a fraction of vehicles being driven with diesel. In Brazil,
all diesle is blended with approximatly 5% of biodiesel, then it is named as
B5. The categorization of the names in LCV is similar with PC.

```{r complcv, echo = FALSE}
  Vehicles = c("LCV_E25", "LCV_FE25", "LCV_FE100", "LCV_E100", "LCV_B5")
  Composition = c("39.13%","15.21%","25.90%", "1.18%","18.58%")
df <- data.frame(Vehicles, Composition)
knitr::kable(df, 
  caption = 'Vehicular composition of LCV for applied in this book',
  booktabs = TRUE,
  align = "c"
)
```


```{r eval = FALSE}
LCV_E25 <- age_ldv(x = net$ldv,name = "LCV_E25",
                   k = 10/100*39.13)
LCV_FE25 <- age_ldv(x = net$ldv, name = "LCV_FE25", agemax = 13,
                    k = 10/100*15.21)
LCV_FE100 <- age_ldv(x = net$ldv, name = "LCV_FE100", agemax = 13,
                     k = 10/100*25.90)
LCV_E100 <- age_ldv(x = net$ldv, name = "LCV_E100", agemin = 9,
                    k = 10/100*1.18)
LCV_B5 <- age_ldv(x = net$ldv, name = "LCV_B5",
                  k = 10/100*18.58)
saveRDS(LCV_E25, file = "veh/LCV_E25.rds")
saveRDS(LCV_FE25, file = "veh/LCV_FE25.rds")
saveRDS(LCV_FE100, file = "veh/LCV_FE100.rds")
saveRDS(LCV_E100, file = "veh/LCV_E100.rds")
saveRDS(LCV_B5, file = "veh/LCV_B5.rds")
```


**Heavy Good Vehicles**

HGV uses diesel which is blended with approximatly 5% of biodiesel from sugar-
cane. 


```{r comphgv, echo=FALSE}

Vehicles = c("SLT", "LT", "MT", "SHT", "HT")
Composition = c("8.38%","25.50%", "15.28%", "24.98%", "25.85%")
df <- data.frame(Vehicles, Composition)
knitr::kable(df, 
caption = 'Vehicular composition of HGV for applied in this book',
booktabs = TRUE,
align = "c"
)
```



```{r eval = FALSE}
HGV_SLT_B5 <- age_hdv(x = net$hdv,name = "HGV_SLT_B5",
                      k = 75/100*8.38)
HGV_LT_B5 <- age_hdv(x = net$hdv, name = "HGV_SLT_B5",
                     k = 75/100*25.50)
HGV_MT_B5 <- age_hdv(x = net$hdv, name = "HGV_SLT_B5",
                     k = 75/100*15.28)
HGV_SHT_B5 <- age_hdv(x = net$hdv, name = "HGV_SLT_B5", 
                      k = 75/100*24.98)
HGV_HT_B5 <- age_hdv(x = net$hdv, name = "HGV_SLT_B5",
                     k = 75/100*25.85)
saveRDS(HGV_LT_B5, file = "veh/HGV_SLT_B5.rds")
saveRDS(HGV_SLT_B5, file = "veh/HGV_LT_B5.rds")
saveRDS(HGV_MT_B5, file = "veh/HGV_MT_B5.rds")
saveRDS(HGV_SHT_B5, file = "veh/HGV_SHT_B5.rds")
saveRDS(HGV_HT_B5, file = "veh/HGV_HT_B5.rds")
```



**Buses**

In Brazil there are many type of buses, but in this book we are focussing in the
most abundandt: Urban Buses, Small Urban Buses and Motoreway Buses ro Coach.
According to the Secretary or Urban Mobility of São Paulo
(Sptrans, http://www.sptrans.com.br/), the fleet has an average age of use of
5 years and 5 monthts. To achieve this average age, the agemax of this vehicles
is 10 years of use.

```{r compbus, echo=FALSE}

  Vehicles = c("UB", "SUB", "MB")
  Composition = c("77.43%", "9.07%", "13.5%")
df <- data.frame(Vehicles, Composition)
knitr::kable(df, 
  caption = 'Vehicular composition of BUS for applied in this book',
  booktabs = TRUE,
  align = "c"
)
```




```{r eval = FALSE}
UB_B5 <- age_hdv(x = net$ldv,name = "UB_B5", agemax = 10,
                 k = 25/100*77.43)
SUB_B5 <- age_hdv(x = net$ldv, name = "SUB_B5", 
                  k = 25/100*9.07)
MB_B5 <- age_hdv(x = net$ldv, name = "MB_B5",
                 k = 25/100*13.5)
saveRDS(HGV_LT_B5, file = "veh/HGV_SLT_B5.rds")
saveRDS(HGV_SLT_B5, file = "veh/HGV_LT_B5.rds")
saveRDS(HGV_MT_B5, file = "veh/HGV_MT_B5.rds")
```



**Motorcycles**

The vehicular composition consisted in Motorcycles using E25, and recnetly,
in year 2010, entered into the market flex motorcycles, which can use
gasoline E25 or ethano E100. Therefore, the oldest flex MC is 6 years old.


```{r compmc, echo=FALSE}

  Vehicles = c("MC_150_E25","MC_150_500_E25", "MC_500_E25",
             "MC_150_FE25","MC_150_500_FE25", "MC_500_FE25",
             "MC_150_FE100","MC_150_500_FE100", "MC_500_FE100")
  Composition = c("72.97%", "11.28%", "3.15%",
            "3.93%", "0.57%", "0.16%",
            "6.69%", "0.98%", "0.27%")
df <- data.frame(Vehicles, Composition)
knitr::kable(df, 
  caption = 'Vehicular composition of MC for applied in this book',
  booktabs = TRUE,
  align = "c"
)
```



```{r eval = FALSE}
MC_150_E25 <- age_hdv(x = net$ldv,name = "MC_150_E25",
                      k = 15/100*72.97)
MC_150_500_E25 <- age_hdv(x = net$ldv, name = "MC_150_500_E25", 
                          k = 15/100*11.28)
MC_500_E25 <- age_hdv(x = net$ldv, name = "MC_500_E25",
                      k = 15/100*3.15)

MC_150_FE25 <- age_hdv(x = net$ldv,name = "MC_150_FE25", agemax = 6,
                       k = 15/100*3.93)
MC_150_500_FE25 <- age_hdv(x = net$ldv, name = "MC_150_500_FE25", agemax = 6, k = 15/100*0.57)
MC_500_FE25 <- age_hdv(x = net$ldv, name = "MC_500_FE25", agemax = 6,
                       k = 15/100*0.16)

MC_150_FE100 <- age_hdv(x = net$ldv,name = "MC_150_FE100", agemax = 6,
                        k = 15/100*6.69)
MC_150_500_FE100 <- age_hdv(x = net$ldv, name = "MC_150_500_FE100", agemax = 6,
                            k = 15/100*0.98)
MC_500_FE100 <- age_hdv(x = net$ldv, name = "MC_500_FE100", agemax = 6,
                        k = 15/100*0.27)

saveRDS(MC_150_E25, file = "veh/MC_150_E25.rds")
saveRDS(MC_150_500_E25, file = "veh/MC_150_500_E25.rds")
saveRDS(MC_500_E25, file = "veh/MC_500_E25.rds")

saveRDS(MC_150_FE25, file = "veh/MC_150_FE25.rds")
saveRDS(MC_150_500_FE25, file = "veh/MC_150_500_FE25.rds")
saveRDS(MC_500_FE25, file = "veh/MC_500_FE25.rds")

saveRDS(MC_150_FE100, file = "veh/MC_150_FE100.rds")
saveRDS(MC_150_500_FE100, file = "veh/MC_150_500_FE100.rds")
saveRDS(MC_500_FE100, file = "veh/MC_500_FE100.rds")

```




# Appendix B {-}
